\section{Uncertainty}

\subsection{For unitary evolutions}

In a continuous-time Page-Wootters universe,
one is tempted to observe that in $\hilb{H}_T$,
in the time representation, it is $E \eqbydef -\hbar\hat{\Omega} \repr i\hbar\partial_{t}$,
in analogy with $\hat{p} \repr -i\hbar\partial_{q}$ in $\hilb{H}_S$,
therefore a time-energy uncertainty relation can be derived
with a proof that is formally identical to the well known
position-momentum uncertainty relation. However, the analogy of
$\hat{\Omega}$ with momentum $\hat{p}$,
and thus of $\hat{T}$ with position $\hat{q}$,
would be correct only if there has a probability distribution \emph{in time},
which is not the case: unitary evolution means a particle exists with
certainty at all times. Therefore, a proof of a time-energy uncertainty
relation based on the formal analogy with position and momentum
(and, of course, the fact that time is finally an operator bypassing the Pauli objection)
can only apply to \emph{events}, in the sense of \eqref{eq:pwphi}.

One may argue that the unitary evolution
\[
  \dket{\Psi} = \int \dd{t} \ket{t}_T \ox \ket{\psi(t)}_S
\]
is a special case of
\[
  \dket{\Phi} = \int \dd{t} \phi(t) \ket{t}_T \ox \ket{\psi(t)}_S
\]
with $\phi(t)=1 \; \forall t$ (i.e. an event that is indefinitely spread in time).

As a function of $t$, $\ket{\psi(t)}_S$ will have a Fourier transform
$\ket{\tilde{\psi}(\omega)}$
so that it will also be
\[
  \dket{\Psi} = \int \dd \omega \ket{\omega}_T \ox \ket{\tilde{\psi}(\omega)}_S \, \text{,}
\]
where $\ket{\tilde{\psi}(\omega)}_S$ is not necessarily normalized for each $\omega$
(and actually it's zero if $\hbar\omega$ does not belong to the energy spectrum of the system).
An extreme case is when $\dket{\Psi}$ encodes what in ordinary quantum mechanics would be the evolution
of an energy eigenstate $\ket{\epsilon}_S$: in such case
$\ket{\tilde{\psi}(\omega)}_S = \delta(\omega - \frac{\epsilon}{\hbar})\ket{\epsilon}_S$
and it is $\sigma_{T} = \infty$ and $\sigma_{\Omega} = 0$.

All other cases of unitary evolution correspond to $\sigma_{T}$
sill being infinite, but $\sigma_{\Omega}$ being finite, therefore
$\sigma_{T}\sigma_{\Omega} = \infty$, which is not particularly interesting:
any possible uncertainty relation would be valid, and $\dket{\Psi}$
is certainly \emph{not} a minimum uncertainty state\dots 

\subsection{For \emph{events}}

Towards proving a time-energy uncertainty relation in $\hilb{H}_T$
based on formal analogy with position and momentum in $\hilb{H}_S$,
a normalized element $\dket{\Phi}$ of $\hilb{H}_T\ox\hilb{H}_S$ 
is not not affected by the difficulties mentioned above.
Still, it is not generally
a separable state,
and thus
the problem of time and energy relation cannot be reduced to that of
a pure state in $\hilb{H}_T$.
It is in general an entangled state in the product space.
Thus the spatial degrees of freedom should be \emph{traced out},
before deriving a time-energy relation.

With a slight change of notation we express:
\begin{equation}
  \dket{\Phi} =
    \int \dd{t} \ket{t}_T \ox \ket{\phi(t)}_S =
    \int \dd\omega \ket{\omega}_T \ox \ket{\tilde{\phi}(\omega)}_S \, \text{,}
\end{equation}
with $\ket{\tilde{\phi}(\omega)}_S$ being the Fourier transform
of $\ket{\phi(t)}_S$,
neither kets being necessarily normalized, which is by the way necessary
to satisfy the normalization in time (frequency)
\[
  \int \dd{t} \braket{\phi(t)}_S =
    \int \dd{\omega} \braket{\tilde{\phi}(\omega)}_S =
    \dbradket{\Phi}{\Phi}_{T \ox S} = 
    1
\]
while of course being $\setof{\ket{t}_T}$ and $\setof{\ket{\omega}_T}$
orthonormal.

Next we use \eqref{eq:density_A_expand}: replace the discrete sum with an integral,
set $i = \mu$ as this is a Schmidt decomposition, and then $j = t$
(or $j = \omega$ respectively).
The reduced density operator can so be computed
as partial trace:
\begin{align}
  \hat{\rho}^T = \Tr_S\qty(\dketdbra{\Phi}{\Phi}) &= \int \dd t \norm{\phi(t)}^2_S \ketbra{t}{t}
    \, \text{,}
  \\ 
  \hat{\rho}^T = \Tr_S\qty(\dketdbra{\Phi}{\Phi}) &= \int \dd \omega \norm{\tilde{\phi}(\omega)}^2_S \ketbra{\omega}{\omega}
    \,\text{.} 
\end{align}
Here the probabilty distributions $\norm{\phi}^2$ and $\norm{\tilde{\phi}}^2$
are ``classical'' in the sense of a mixed state.

The relation $\sigma_T\sigma_{\hbar\Omega} = \hbar \sigma_{\phi} \sigma_{\tilde{\phi}} \geq \frac{\hbar}{2}$
can then be derived from the properties of the Fourier transform.\footnote{
  Fourier transform of a
  vector-valued function
  (vectors in $\hilb{H}_S$),
  similar to what found in \cite{Maccone:Pauli}.
}

\if{todo}
\subsection{TODO}
\cite{Maccone:Pauli} \S IV.  UNBOUNDED-ENERGY CLOCKS?
\fi