\documentclass{book}

\input{tex/lib}

\author{Guido De Rosa \\ \small\tt{gderosa@umail.ucc.ie}}
\title{Time as a quantum observable: from Pauli objection to Page and Wootters model --- improvements and applications}

\usepackage[safeinputenc]{biblatex}
\addbibresource{biblio.bib}

\begin{document}

\maketitle

\tableofcontents

\chapter*{Abstract}
A proof is detailed of the (in)famous Pauli's ``theorem''~\cite{PauliFootnote}
on the impossibility of a time observable in quantum mechanics. Some possible
approaches towards overcoming Pauli's objections are reviewd and expanded as well

\chapter{Motivation and experiments}
\input{tex/intro/intro}

\chapter{Pauli's objection --- TODO: make this Historical Introduction?}
\input{tex/pauli/intro}
\input{tex/pauli/proof}

\chapter{Approaches}
\input{tex/approaches/approaches}

\chapter{Entanglement, decoherence, and measurement}
The environment is watching, Haroce-Raimond, from ``Exploring the quantum'', ch.4

Environment-Induced Decoherence and the Transition from Quantum to Classical

Partial trace, which is not a number, check wikipedia $\rightarrow$ consistent histories.
J. P. Paz W. H. Zurek

Von Neumann and Shanon entropy.

MIKIO NAKAHARA: QUANTUM COMPUTING: AN OVERVIEW

\section{Maximally mixed state of $\ket{e_1}$ and $\ket{e_1}$
Vs their maximal coherent superposition: density matrices}

See Nakahara notes, example II.2.

$$
\frac{1}{2}\begin{bmatrix}
  1 &0  \\
  0 &1  \\
\end{bmatrix}
\text{ Vs }
\frac{1}{2}\begin{bmatrix}
  1 &1  \\
  1 &1  \\
\end{bmatrix}
$$

The off-diagonal terms in the coherent superposition case indicate
``quantumness''. (explain)

\begin{remark}
  The density matrix of a pure state is either diag(000\dots1\dots000) or non-diagonal.

  It can be easily proven reasoning on eigenvalues\dots

  So, except the special/trivial case above, a pure state must have off-diagonal terms.
\end{remark}

Measurement destroys off-diagonal terms and turns a pure state into a mixed one.

Off-diagonal terms in the density matrix indicates ``quantum purity''.
(Or internal Entanglement as well, if the system being detected was bipartite itself?)

In some sense, measurement turned a q-statistical distribution of values for the observable,
into a classical probability distribution.

Recommended read: \cite{Zurek_Decoherence, Zurek_Decoherence2}.

\chapter{Critical review of Page and Wootters mechanism and proposed improvements}
\chaptermark{Critical review of Page and Wootters \dots}
\input{tex/pw/review}
\input{tex/pw/lg}

\chapter{Reconciling the Detector Model with the Page and Wootters mechanism?}
\section{Can a POVM on a system, if then we see it as part of a bipartite one,
equivalent to a PVM on the other, entangled, system?}
This will show an equivalence of models based on POVM with the Page and Wootters...?

Not exactly, but a POVM on a subsystems corresponds to a PVM on the tensor product space.

From \cite{PreskillNotes}, Ch.3 
\begin{quotation}
We have seen that
a pure state of the bipartite system AB may behave like a mixed state
when we observe subsystem A alone, and that an orthogonal measurement
of the bipartite system can realize a (nonorthogonal) POVM on A alone.
\end{quotation}

The same chapter talks about quantum operations, quantum channels and Kraus opertators.

We might want to look at exponential decay from \url{https://arxiv.org/abs/1704.07236},
then compare with exponential decay with P and W using Lloyd Giovannetti and Maccone (ref).

\subsection{4-partite universe?}
\begin{itemize}
  \item{The system being measured/detected}
  \item{The Ruschhaupt detector --- which does not measure time, but whose detection happens at a certain time}
  \item{The Page and Wootters clock, entangled with the system and/or the detector}
  \item{The rest of the Universe, aka the Environment, aka the Termal Bath or Reservoir}
\end{itemize}

Can any of the above be identified? If the lab is isolated enough,
the detector is the only macro object and can act as a Universe/bath/environment/reservoir\dots?

\chapter{When Time Crystals come into play}
\input{tex/crystals/crystals}

\chapter{Thermal time hypothesis}
\input{tex/therm/connes-rovelli}

\chapter{Gravitation and High Energy}
\input{tex/grav-he/feynm}
\input{tex/grav-he/grav}

\chapter{Extras}
\input{tex/misc/misc.tex}

\chapter{Pedagogical remarks (TODO: move to Intro? move to Conclusions?)}
A computation/information approach for a better understanding of quantum
physics.\footnote{
  Namely, Preskill notes \cite{PreskillNotes} as a better way of looking at QM,
  especially \textbf{measurement}.
}

\appendix
\chapter{Commutator properties}
\input{tex/appendix/commutator}
\chapter{Density operators, projectors and measurement}
\input{tex/appendix/density}

\printbibliography[title=References]

\end{document}