\section{Unitary evolution \emph{versus} absorption}

A vector $\dket{\Psi}$ in $\hilb{H}_T \ox \hilb{H}_S$,
satisfying \eqref{eq:pwHamiltonian} and \eqref{eq:Wheeler-DeWitt},
encodes the whole (unitary) time evolution of a system.
\begin{equation}\label{eq:pwexpansion}
  \dket{\Psi} =
    \int \dd{t} \ket{t}_T \ox \ket{\psi(t)}_S =
    \int \dd{t}\dd[3]{\vec{r}} \ \Psi(t; \vec{r}) \; \ket{t}_T \ox \ket{\vec{r}}_S
    \,  \text{.}
\end{equation}
We know $\setof{\ket{t}_T \ox \ket{\vec{r}}_S}$ is an othonormal basis of $\hilb{H}_T \ox \hilb{H}_S$, therefore
\begin{equation}
  \norm{\dket{\Psi}}^2 =
    \int \dd{t}\dd[3]{\vec{r}} \ \abs{\Psi(t; \vec{r})}^2 =
    \int \dd{t} \int \dd[3]{\vec{r}} \ \abs{\Psi(t; \vec{r})}^2 =
    \int \dd{t} 1 \rightarrow +\infty
    \,  \text{,}
\end{equation}
which means that such $\dket{\Psi}$ is an \term{improper} vector of $\hilb{H}_T \ox \hilb{H}_S$.

Proper (i.e. normalizable) states are described in \cite{Lloyd:Time} as well, by replacing (or generalizing)
the \eqref{eq:pwexpansion} with
\begin{equation}\label{eq:pwphi}
  \dket{\Phi} =
    \int \dd{t} \phi(t) \ket{t}_T \ox \ket{\psi(t)}_S \, \text{.}
\end{equation}
If the function $\phi \in \mathcal{L}^2(\mathbb{R})$,
then $\dket{\Psi}$ is a proper element of the product space,
and $\norm{\dket{\Psi}}^2 = \int \dd{t} \abs{\phi(t)}^2$.

The case of non-normalizable $\dket{\Psi}$ in \eqref{eq:pwphi},
with normalized $\ket{\psi(t)}_S$ $\forall t \in \mathbb{R}$,
describes the unitary evolution, as seen throughout Chapter \ref{ch:pw}.
As observed in \cite{Maccone:QGR},
``%
  Quantum mechanics is formulated in terms of \emph{systems},
  typically limited in space but infinitely extended in time%
''.
If the state vector is \emph{conditioned} at a particular time $t$,
it holds $\norm{_{T}\bradket{t}{\Psi}}_S = \norm{\ket{\psi(t)}}_S = 1$,
meaning that, at each $t$,
\emph{the particle must certainly be in some (one) point in space}.

A normalized $\dket{\Psi}$ in the whole $\hilb{H}_T \ox \hilb{H}_S$,
instead,
can be interpreted as a total probability of~$1$ in both space and time combined.
It can be interpreted as describing an \term{event},
in that it
must certainly be in some point in space
\emph{and} time (in terms of outcome of an idealized measurement\footnote{
  Measurement in quantum mechanics requires the concepts
  of state of the system
  (and measurement apparatus)
  \emph{before} and \emph{after} the measurement, therefore the existence
  of an external time (external with respect of the Hilbert space of states).
  This logically contradicts the foundation of the Page--Wooters model if
  time itself is measured as a quantum observable. $\dket{\Psi} \in \pwspace$
  embeds the whole history of a system and therefore cannot have a
  ``before'' neither an ``after'' the measurement, ``when'' it collapses
  into an eigenstate of time. The apparent contradicion is resolved
  stressing that probability amplitues are intended in the sense of
  \emph{conditional} probabilities e.g. \emph{provided that the particle
  is in position} $x \in X$ (or the detector clicks)
  what is the probability density amplitude of time being (the clock showing) $t$?
  Consistently, the Bayes rule \parencite[\S 8.7]{StatBook} is invoked in the following sections
  and references.
}).
A ``typical'' example of localized ``event wave packet'' would be
therefore represented by
a \emph{4-dimensional} gaussian wave function,
in analogy to well known examples of purely spatial gaussian states
in quantum mechanics and quantum optics.

The detector absoption model in \cite{RuschhauptAbsorption} is not originally
based on the notion of quantum relational time. It's ``ordinary quantum mechanics''
in the sense that states are in $\hilb{H}_S$ with an external parameter time;
but, contrary to quantum mechanics, the potential can be \emph{complex},
causing the norm of the state vector to not be conserved (in fact, vanishing with time).
In other terms, the evolution is not unitary, even for a pure state.
Specifically, the Hamiltonian is corrected by an anti-hermitian term
that models the detector.

A non hermitian Hamiltonian is justified as a computation method
to simplify the study of some open systems: the evolution of mixed
states is derived without explicit reference to density operators
or master equations, but resolving equations that are formally
identical to those of pure states,
i.e. in terms of
Schr{\"o}dinger equations and wave functions,
with the non-hermitian term in the Hamiltonian
to account for the non-unitarity of the evolution.
%
This is treated extensively in:
  \cite[Ch. 6]{TQM2};
  \cite{Wave-function_approach};
  \cite{HowToResetAnAtom};
  \cite{TheQuantumJumpApproach};
  \cite[\S 8.5.2 ``The `quantum jump' approach to damping: The wave function Monte Carlo approach'']{ScullyZubairy};
  \cite[\S 6.7.1 ``Simulating Quantum Trajectories'']{WallsMilburn}.

% In comparing the results in (or attempting to draw some common conclusion from)
% \cite{Lloyd:Time} and \cite{RuschhauptAbsorption}
% ---or matching the \eqref{eq:pwphi} and \eqref{eq:pwf} if one will---
% different approaches are conceivable:
% \begin{enumerate}
%   \item
%     Setting
%     \begin{equation}
%       \phi = f \, \text{,}
%     \end{equation}
%     after checking that relevant results in \cite{Lloyd:Time} are valid in more general cases than
%     $\phi \in \mathcal{L}^2(\mathbb{R})$
%   \item
%     Honoring the ``event'' semantics proposed above, and the notion of \term{absorption event}:
%     something that has its peak probability (density)
%     when the norm of $\ket{\psi(t)}$ decreases most rapidly; therefore if we can assume
%     $\dv{\abs{f}^2}{t} \leq 0 \; \forall t$, the probability amplitude density
%     (or the wavefunction of absorption event) will be a $\phi$ such that
%     \begin{equation}\label{eq:absorptionEvent}
%       \abs{\phi(t)}^2 = -\dv{\abs{f}^2}{t}
%     \end{equation} 
% \end{enumerate}

% The requirement, implicit  in \eqref{eq:absorptionEvent},
% that $f$ never locally increases in modulus
% is not restrictive, or it's not more restrictive than simply stating
% that the detector, at most, absorbs, but never \emph{emits}
% (in the model of \cite{RuschhauptAbsorption}).
