\section{Detector models}

As we have seen in the previous sections,
proper elements of $\pwspace$,
localized in both space and time,
can be interpreted as ``events''
(as opposed to full history vectors, which instead describe the unitary evolution
of a quantum system \emph{at all times}).

The arrival of a particle at a detector
is a kind of event
of particular interest from the perspective of time
in quantum mechanics, for several reasons.

The first reason
is the direct connection with experiments,
``given the appalling evidence that time is also a random variable in the laboratories''
\parencite{TQM2:Detector};
and because,
``{[as]} a matter of fact, a number of time observables are already routinely measured in laboratories,
for example arrival times in time-of-flight experiments,
but the theoretical foundation of these measurements is still being discussed''
\parencite[Preface to the First Ed.]{TQM1}.

The second reason is the existence of studies
about detector models which also investigate
time-of-arrival as a quantum observable.
In particular we will consider \cite{RuschhauptAbsorption},
which in turn traces back its basic ideas from the seminal work by Allcock
in the 1960s \parencite{Allcock-1, Allcock-2, Allcock-3},
while a detailed contemporary formulation can be found in
\cite{TQM2:Detector}.
None of these works are explicitly
based on Page--Wootters relational model ---of which, to our knowledge,
there are no working examples of application to
time-of-detection problems in the current literature.
Section \ref{sec:absorption+pw} and the remainder of the chapter
will be devoted to bridging such gap
by implementing such application
and comparing
the results from the different models.
Some emphasis will be given to
\emph{discrete} relational time
using the techniques developed in Chapter \ref{ch:pw}.

\subsection{The papers by Allcock and following developments}

With three papers dated 1969 \parencite{Allcock-1, Allcock-2, Allcock-3}, G. R. Allcock
introduced a time-of-arrival quantum observable,
along with a first formulation of a detector model based on a non-Hermitian
Hamiltonian (by including a \term{complex potential}).
See in particular \cite[\S II-IV]{Allcock-2}, where an anti-Hermitian term
$-iV_0\theta(x)$ is introduced in the Hamiltonian,
to model an absorber aim at detecting
the arrival of a particle in the region $x>0$
(here $\theta$ is the Heaviside \term{step function}).
The particle state evolves
in non-unitary manner, with a transfer of probability
from an ``incident channel''
into ``orthogonal and time-labelled measurement channels''.
Unfortunately Allcock came to negative conclusions regarding this approach.
Among other objections, he noticed that when $V_0$ is large the particle is not absorbed but reflected;
while, when it's small (i.e. the detector has a low absorption rate), the particle is eventually absorbed but
with an indetermination $\delta_T \sim \frac{1}{2}V_{0}^{-1}$
that becomes, in such case, impractically large.
Also, Allcock did not directly tackle the Pauli objection (see Section \ref{proof})
but rather maintained that argument.

The Pauli objection can be overcome by the Page--Wootters mechanism
in that a time observable is defined in a separate Hilbert space
than the one where the Hamiltonian is defined,
thus dropping a fundamental assumption of the Pauli argument.
One of the goals of this chapter
is indeed combining the two models. Another route is based on POVM
(a more general, ``unsharp'' measurement, as opposed to projective measurement; a concept we have reviewed in Chapter~\ref{ch:decohere}).
Quantum time measurement can be formulated as a POVM
(see ```Standard' Quantum-Mechanical Approach to Times of Arrival'', \cite{TQM1:Standard}).
This ``standard'' approach does not require an extra Hilbert space \emph{for the clock}.

Nonetheless, we know that a projective measurement on a bipartite system
can be regarded as a POVM
if we look at
on one part only (e.g. \cite{Paris2012}),
thus the clock space of the Page--Wootters mechanism, $\hilb{H}_T$,
can be regarded as a purification space \parencite{Paris2012} for the ``standard''
Hilbert space where the time-of-arrival POVM is defined,
and an interesting line of further research would be a more formal
study of the logical connection between the two approaches, based on this principle.
Both models respond to the issues raised by Pauli
by giving up
the pursuit of
a \emph{projective} measurement in the \emph{standard} Hilbert space,
and rather ``opening'' that Hilbert space in some sense.

The limitations of complex potentials either causing reflection
or not absorbing sufficiently have been resolved
in the following literature
by noting that the results by Allcock were not general, and
potentials can be constructed that absorb the whole wave packet
and avoid reflection \parencite{Muga_TOAQM, Muga_CompositeAbsPot, ComplexAbsPot}.

\subsection{How a complex potential emerges}

Among the methods to define a time-of-arrival quantum observable,
the detector model by Allcock
---including the following enhancements---
is an \emph{operational} one,
in that it models a measurement procedure \parencite[\S 9]{Leavens_TOA}
rather than focusing on constructing an operator that satisfies certain requirements
(like the Kijowski distribution, see for example \cite[\S 8]{Leavens_TOA})
or on building an extended, general framework
like the Page and Wootters model, at least in its original formulation.

The introduction of complex potentials, a non-Hermitian Hamiltonian,
the non-unitary evolution (of a pure state!) may appear artificial to some,
and in contradiction with the fundamental rules of quantum mechanics.

\citereset
In fact, as pointed out well in \cite{Halliwell_Irreversible},
an \emph{effective} complex potential and the consequent loss of normalization
only emerge after ``tracing out'' a portion of the whole system,
which includes the particle, the detector and the environment altogether.
Such whole tripartite system \emph{is} closed, in a pure state, and evolves unitarily.
The particle wavefunction is also an ``effective'' one derived from its density operator.

Not surprisingly, absorption is an \term{irreverersible} process.
While a complex potential can be \emph{postulated} in order to obtain
phenomenological laws governing arrival times and detectors,
it can also be \emph{derived} from first principles and within the established framework
of open quantum systems, namely the master equation
for an irreversible detector.
The core ideas behind this derivation
can be ---very briefly--- summarized as follows.

The detector is modeled by a two-level system with $\ket{1}$ being the state of no-detection
and $\ket{0}$ the state of detection.
We also introduce the raising and lowering operators $\sigma_{+}=\ketbra{1}{0}$, $\sigma_{-}=\ketbra{0}{1}$.
The Hamiltonian of the detector is such to have $\ket{0}$
at a lower energy, so the detector ``decays'' when it ``clicks''. It's an irreversible transition because
of the coupling with the environment.
The Hamiltonian encompassing the particle, the detector, the environment the interaction reads
\begin{equation}
  H = H_s + H_d + H_{E} + V(x)  H_{dE} \,\text{,}
\end{equation}
with $H_s$ being the Hamiltonian of a free particle and
\begin{align}
  H_d     &= \frac{1}{2}\hbar\omega \qty(\ketbra{1} - \ketbra{0}) \\
  H_{E}   &= \sum_n \hbar \omega_n a_n^{\dag} a_n \\
  H_{dE}  &= \sum_n \hbar \left( \kappa^*_n \sigma_{-} a_n^{\dag} + \kappa_n \sigma_{+} a_n \right) \\
  V(x)    &= \theta(-x) \,\text{,}
\end{align}
where $a$ and $a^{\dagger}$ are the creation and annihilation operator for the electromagnetic field,
which constitutes the ``environment''; $V(x)$ is chosen to be a step function
i.e. the simplest function that makes the detector respond
when the particle reaches the region ($x<0$);
and the coupling constants $\kappa_{n}$ are to be intended in the same sense of the Jaynes--Cummings model
---in fact, the expressions of $H_d$ and $H_E$ also recall it
(\cite[\S 10.2]{WallsMilburn}, \cite{JCM} and many others).
This model essentially describes the detector as a Jaynes-Cummings atom
with the peculiarity that its coupling with the environment
is only activated by the position distribution
of another particle.

\subsubsection{TODO cite}
\cite{Damborenea}, or better \cite{Damborenea_atomic}.

\subsection{Then I would go straight to Kiukas article}

Duh.

\subsection*{TODO??}

Use the detector chapter Ruschhaupt book for both getting more references above and description following.

Maybe\dots? \cite{ProbCurrent}, \cite{Ruschhaupt_QMoT}.

``Book vol. 2'' \cite{TQM2:Detector}.

Cite \cite{Muga_ArrTimeOpNormal, Sudarshan_Zeno, Echanobe, Savvidou-1, Savvidou-2}.

