\section{Quantum and the Environment: from Quantum to Classical}

Placeholder text.

\section{TODO}

Measurement destroys off-diagonal terms and turns a pure state into a mixed one.

Off-diagonal terms in the density matrix indicates ``quantum purity''.
(Or internal Entanglement as well, if the system being detected was bipartite itself?)

In some sense, measurement turned a q-statistical distribution of values for the observable,
into a classical probability distribution.

Recommended read: \cite{Zurek_Decoherence, Zurek_Fundamentals}.

Zurek above: DeWitt, Everett, gell_Mann, hartl, Many Worlds, consistent/decoherent historis:
idea: Lagrangian over a history? Principle of least action?

Zuerk: Von Neumann: measurement apparatus as quantum, not classical (as opposed to Bohr).
TODO: use this to expand Historical Intro as ``the book'' is fairly brief in relation to
Von Neumann.

Nakahara, Ch. 9 book, ``Tracing out the extra degrees of freedom makes it impossible to invert a quantum operation.''

The environment is watching, from ``Exploring the quantum''
\parencite[Ch. 4]{Haroche_Exploring}. Start with subsect. 2.4.1.

2.5.4 The quantum–classical boundary
Decoherence models versus Copenhagen interpretation.

Environment-Induced Decoherence and the Transition from Quantum to Classical,
\cite{Zurek_Fundamentals}.

Partial trace, which is not a number, check wikipedia $\rightarrow$ consistent histories.
J. P. Paz W. H. Zurek

Von Neumann and Shanon entropy.

MIKIO NAKAHARA: QUANTUM COMPUTING: AN OVERVIEW

See also \cite{Schlosshauer_Decoherence}.

Vedral, Brukner (Oreshkov\dots)
Necessary and sufficient condition for non-zero quantum discord
\url{https://arxiv.org/pdf/1004.0190.pdf},
cites another paper by Zurek \url{https://arxiv.org/pdf/quant-ph/0105072.pdf}.
