This work relates to the \emph{problem of time} in quantum physics,
% \footnote{
%   See, for a review, \cite{TQM1, TQM2}.
% }
from Wolfgang Pauli's argument
on the impossibility of defining a time observable, %\parencite{PauliFootnote},
to a number of models aimed at overcoming it,
and in particular \emph{relational} models
developed during the last decades.

According to the theoretical framework first proposed by Page and Wootters
(and later improved by Lloyd, Giovannetti and Maccone),
time and unitary evolution only emerge in
terms of \emph{entanglement} between non-interacting subsystems
---one of which acts as a ``clock''
of an otherwise stationary universe. % \parencite{Marletto:Evolution}.

Historically, Pauli's ``no-go'' argument was simply a brief observation in a footnote
of his \textit{General Principles of Quantum Mechanics},
with no rigorous or complete proof. Recent works, outlining such proof, are
reviewed, while providing some additional details.

One of the aims of the thesis
is contributing towards bridging a gap between the theoretical
and experimental literature on this matter.
To this end, original examples and applications are developed on top of existing theoretical results.
Conversely,
a critical review
of recent experiments is carried out.
Numerical computation is also extensively employed in order to compare
predictions from different models. A particular attention is given
to finite-level clocks,
thus extending the Page--Wootters formalism ---which is based on continuous time---
to cases that are easier to implement numerically;
have no mathematical difficulties in terms of convergence;
and, in principle, can be realized experimentally through existing quantum technologies.