{
  % [final only]

  I've been working towards this Master's degree
  while continuing a full-time career
  as an IT professional and software engineer.
  I want to thank my
  colleagues and friends for their encouragement
  and intellectual curiosity
  about my endeavour.
  I also want to thank those managers
  who have, in various ways, supported or facilitated it.

  I want to thank my family, friends, adored ones back in the continent,
  who I am not seeing ---in person--- for a while during this pandemic;
  nonetheless they have been always present.
  
  Thanks to all my fellow postgraduate students,
  as well as research staff,
  at the ``Shortcuts to Adiabaticity'' group in UCC
  (and one senior collaborator in particular!).
  Thanks to my {supervisor}, for his immense patience,
  the advice against numerous ``traps'' along the way, and many, many other qualities.
  Also thanks to all those involved in the teaching lab, for the experience as a
  first-year demonstrator that,
  if not strictly connected to the topic of this thesis,
  does remind me where all modern physics ultimately comes from.

  I am grateful to
  Nicole,
  Paul,
  Irfan,
  Antonello,
  Andreas,
  Mirek,
  Paola,
  Marco,
  Jalal,
  Clarissa,
  Mark,
  Dave,
  Sergio,
  Rocco,
  Rosario,
  Trevor,
  Noel,
  Chris,
  Tiziano,
  Sandra,
  Anthony,
  Tom,
  Stephen,
  Lukas,
  Antonio,
  Elena,
  Cian,
  Nasir,
  Tamara,
  Deborah,
  KC,
  Nico,
  and Margherita.
  In random order, removing duplicates, and certainly missing many ---to whom I apologize.
}
